```{r}
#### Setup
# Load packages
library(blackmarbler)
library(geodata)
library(sf)
library(terra)
library(ggplot2)
library(tidyterra)
library(lubridate)
library(tigris)

#### Define NASA bearer token
bearer <- "eyJ0eXAiOiJKV1QiLCJvcmlnaW4iOiJFYXJ0aGRhdGEgTG9naW4iLCJzaWciOiJlZGxqd3RwdWJrZXlfb3BzIiwiYWxnIjoiUlMyNTYifQ.eyJ0eXBlIjoiVXNlciIsInVpZCI6InNtYWxsbGlnaHQiLCJleHAiOjE3OTI4Nzg5MDEsImlhdCI6MTc4NzY5NDkwMSwiaXNzIjoiaHR0cHM6Ly91cnMuZWFydGhkYXRhLm5hc2EuZ292IiwiaWRlbnRpdHlfcHJvdmlkZXIiOiJlZGxfb3BzIiwiYWNyIjoiZWRsIiwiYXNzdXJhbmNlX2xldmVsIjozfQ.TMWqEtpPA3BPxT5i8xdz816DXpFo1PJdIljZM3YIzBcxzhLkc62sCecxTKTIeMBnb9Dxuqeejn_4ctcBMyWhO7XGdYJn0-a-zGvwg7k7evfhS98n6AJSbXoUCAmPoTS9FjevLEFZf2liJiaDRQ6PafwnMTy3pPmEnWo8Y1ZGgNBH3NoeITogPrQhismWFvjNOW8IDbNLTs6Cxj-7Kd6RMwn3cXfsJicsOtHB-w62B9GsmUyqtVfLmyC1hvPathcq_Ksl546WkMbE_zd8Rr8bNkWpUchZQbk1bheVlnoEvY78K92ja53_mTuir_BqbDyLDhGnwND7P3MgO_jYdOgR8g"

### ROI
# Define region of interest (roi). The roi must be (1) an sf polygon and (2)
# in the WGS84 (epsg:4326) coordinate reference system. Here, we use the
# getData function to load a polygon of Ghana
roi_sf <- gadm(country = "GHA", level=1, path = tempdir()) 
```

```{r}
### Daily data: raster for February 5, 2021
r_20210205 <- bm_raster(roi_sf = roi_sf,
                        product_id = "VNP46A2",
                        date = "2021-02-05",
                        bearer = bearer)

### Monthly data: raster for October 2021
r_202110 <- bm_raster(roi_sf = roi_sf,
                      product_id = "VNP46A3",
                      date = "2021-10-01", # The day is ignored
                      bearer = bearer)

### Annual data: raster for 2021
r_2021 <- bm_raster(roi_sf = roi_sf,
                    product_id = "VNP46A4",
                    date = 2021,
                    bearer = bearer)
```

```{r}
#### Daily data in March 2021
r_daily <- bm_raster(roi_sf = roi_sf,
                     product_id = "VNP46A2",
                     date = seq.Date(from = ymd("2021-03-01"), to = ymd("2021-03-31"), by = "day"),
                     bearer = bearer)

#### Monthly aggregated data in 2021 and 2022
r_monthly <- bm_raster(roi_sf = roi_sf,
                       product_id = "VNP46A3",
                       date = seq.Date(from = ymd("2021-01-01"), to = ymd("2022-12-01"), by = "month"),
                       bearer = bearer)

#### Yearly aggregated data in 2012 and 2021
r_annual <- bm_raster(roi_sf = roi_sf,
                      product_id = "VNP46A4",
                      date = 2012:2021,
                      bearer = bearer)
```

```{r}

#| eval: false
options(tigris_use_cache = TRUE)

roi_hawaii <- counties(state = "HI", cb = TRUE, year = 2026)
```

```{r}
#| eval: false
library(terra)
crs(roi_hawaii)
```

```{r}
#### Extract annual data
ntl_df <- bm_extract(roi_hawaii = roi_hawaii,
                     product_id = "VNP46A4",
                     date = 2012:2022,
                     bearer = bearer)

#### Trends over time
ntl_df |>
  ggplot() +
  geom_col(aes(x = date,
               y = ntl_mean),
           fill = "darkorange") +
  facet_wrap(~NAME_1) +
  labs(x = NULL,
       y = "NTL Luminosity",
       title = "Ghana Admin Level 1: Annual Average Nighttime Lights") +
  scale_x_continuous(labels = seq(2012, 2022, 4),
                     breaks = seq(2012, 2022, 4)) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

Kona Low Storm

second storm(Mar 19-24)

Cloud cover-?

Weather reporting-?

Satelite imagery-?

Wind Direction-?

How was weather data a week before the storm?

Public HECO data-?

Service hours for affected areas?

When did the power cut officially come back on? 

What infrastructure was damaged during the storm? Power lines, transformers, power stations etc. 

Is HECO rerouting power visible?
