---
title: "GIS for Natural Diaster Response"
format: html
editor: visual
---

### Welcome to our first DS 422 notebook

With the recent power outages from Hurricane Lala, let's explore how we can apply GIS skills to natural disaster management.

#### Data

We will start with the [Eagle-I Power Outage Dataset for 2025](https://www.osti.gov/biblio/3012826) from Oak Ridge National Laboratory.

#### Libraries

```{r}
library(tidyverse)
library(here)
library(sf)
library(tidycensus)
library(arrow)
library(tigris)
library(sf)
library(ggiraph)
library(plotly)
```

I downloaded the original dataset from the link above, filtered to only Hawaii outages, and converted the csv to a feather file to shrink the size for ease of sharing and loading

```{r}
eagle_i_hawaii <- read_feather(here("data/hawaii_outages_2025.feather"))
```

#### Your Turn - Exploratory Data Analysis

Take 15 minutes to explore trends in this data, think summary stats, plots, etc..

```{r}
summary(eagle_i_hawaii)
```

```{r}
ggplot(data=eagle_i_hawaii, aes(x = county, y = customers_out))+ 
  geom_col()
```

#### Outages by county

##### First filter out kalawao

```{r}
eagle_i_hawaii <- eagle_i_hawaii |> 
  filter(county!= "Kalawao")
```

##### Next, we can count outages by county

```{r}
customers_out_by_county <- eagle_i_hawaii |> 
  group_by(county) |> 
  summarize(total_customers_out = sum(customers_out, na.rm = TRUE))
```

##### Your turn - Lets make a bar chart with ggplot

```{r}
ggplot(data= customers_out_by_county, aes(x = total_customers_out, y = county)) + geom_col()
```

##### What if we normalize?

```{r}
hi_pop <- get_acs(
  geography = "county",
  state = "HI",
  variables = "B01003_001",  # total population
  year = 2022
)
```

##### Joining our data

```{r}
hi_pop <- hi_pop |>
  mutate(county = gsub(" County, Hawaii", "", NAME))

hi_pop <- hi_pop |>
  select(county, population = estimate)

customers_out_by_county_with_pop <- customers_out_by_county |>
  left_join(hi_pop, by = "county")
```

##### Normalizing

```{r}
customers_out_by_county_with_pop <- customers_out_by_county_with_pop |>
  mutate(outages_per_1000 = (total_customers_out / population) * 1000)
```

##### Now Plot

```{r}
ggplot(customers_out_by_county_with_pop, aes(y = county, x = outages_per_1000)) +
  geom_col(fill = "firebrick") +
  labs(
    title = "Power Outages per 1,000 Residents by County",
    x = "Outages per 1,000 residents",
    y = NULL
  ) +
  theme_minimal()
```

#### Outages over time

```{r}
outages_over_time <- ggplot(data = eagle_i_hawaii, aes (x = run_start_time, y = customers_out)) +
  geom_line(linewidth = 0.2) +
  theme_minimal()
outages_over_time
```

```{r}
ggplotly(outages_over_time)
```

##### Color by county

```{r}
county_outages_over_time <- ggplot(data = eagle_i_hawaii, aes (x = run_start_time, y = customers_out, color = county)) +
  geom_line(linewidth = 0.2) +
  theme_minimal()
county_outages_over_time
```

```{r}
ggplotly(county_outages_over_time)
```

##### Facet by county

```{r}
ggplot(data = eagle_i_hawaii, aes (x = run_start_time, y = customers_out)) +
  geom_line(linewidth = 0.2) +
  facet_wrap(~county, nrow = 4, scales = "free") +
  theme_bw()
```

#### Map Outages

```{r}
library(tigris)
library(sf)
library(ggplot2)
library(dplyr)
library(patchwork)

# Get Hawaii county boundaries
hi_counties_sf <- counties(state = "HI", cb = TRUE, year = 2022) |>
  mutate(county = gsub(" County", "", NAME)) |>
  filter(county != "Kalawao") |>
  left_join(customers_out_by_county_with_pop, by = "county") |> 
  st_crop(xmin = -160.5, xmax = -154.5, ymin = 18.5, ymax = 22.5)

```

```{r}
ggplot(hi_counties_sf) +
  geom_sf(aes(fill = outages_per_1000)) +
  scale_fill_viridis_c(option = "inferno", name = "Outages\nper 1,000") +
  labs(title = "Power Outages per 1,000 Residents by County") +
  theme_void()
```

##### Bonus

```{r}
p1 <- ggplot(hi_counties_sf, aes(
  x = reorder(county, outages_per_1000),
  y = outages_per_1000,
  tooltip = paste0(county, ": ", round(outages_per_1000, 1), " per 1,000"),
  data_id = county,
  fill = outages_per_1000
)) +
  geom_col_interactive() +
  coord_flip() +
  scale_fill_viridis_c(option = "inferno") +
  labs(x = NULL, y = "Outages per 1,000 residents", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "none")

# ------------------------------------------------------------
# 3. Choropleth map (interactive, severity-colored)
# ------------------------------------------------------------

p2 <- ggplot(hi_counties_sf) +
  geom_sf_interactive(
    aes(fill = outages_per_1000,
        tooltip = paste0(county, ": ", round(outages_per_1000, 1), " per 1,000"),
        data_id = county),
    color = "white"
  ) +
  scale_fill_viridis_c(option = "inferno", name = "Outages\nper 1,000") +
  coord_sf() +
  theme_void()

# ------------------------------------------------------------
# 4. Combine and make interactive
# ------------------------------------------------------------

combined_plot <- p1 + p2 + plot_layout(widths = c(1, 1.2))

interactive_plot <- girafe(ggobj = combined_plot, width_svg = 10, height_svg = 5)
interactive_plot <- girafe_options(
  interactive_plot,
  opts_hover(css = "stroke:black;stroke-width:2px;"),
  opts_hover_inv(css = "opacity:0.3;")
)

interactive_plot
```

#### Whats Next?

This seems like a good start to understanding power outages, but is fairly limited with the county level granularity of the data. Another option to measure power outages is utilizing the [blackmarbler](https://worldbank.github.io/blackmarbler/index.html) package.

**BlackMarbleR** is a R package that provides a simple way to use nighttime lights data from NASA’s Black Marble. [Black Marble](https://blackmarble.gsfc.nasa.gov/) is a [NASA Earth Science Data Systems (ESDS)](https://www.earthdata.nasa.gov/) project that provides a product suite of daily, monthly and yearly global [nighttime lights](https://www.earthdata.nasa.gov/topics/human-dimensions/nighttime-lights). This package automates the process of downloading all relevant tiles from the [NASA LAADS DAAC](https://ladsweb.modaps.eosdis.nasa.gov/) to cover a region of interest, converting and mosaicing the raw files (in HDF5 format) to georeferenced rasters.

#### Your Turn

[Follow the steps here](https://worldbank.github.io/blackmarbler/#installation) to generate the first 2 visualizations of nighttime lights in Ghana (map and bar chart). Stop at "Workflow to Update Data". We will tweak this analysis to measure power outages in Honolulu county pre/during/post severe weather events with known outages from the eagle-i dataset
